\section{Statements and Expressions}

\subsection{Exceptions}

Several locations in this document refer to an exception being thrown.
The semantics of an exception being thrown is that the application
halts with a run-time error.  Future versions of this document will
define facilities for propagating and catching exceptions.

\subsection{Statements}


\begin{quote}
\ntermdef{Stmt}

\defspace \nterm{Expr}

\defspace \nterm{VarDecl}

\defspace \nterm{StateValDecl}


\ntermdef{VarDecl}

\defspace \nterm{Specifier} \opt{\nterm{Type}} \nterm{Identifier} = \nterm{Expr}

\ntermdef{StateValDecl}

\defspace \keyw{stateval} \nterm{Identifier} \nterm{StateBinding}


\ntermdef{Specifier}

\defspace \keyw{val}

\defspace \keyw{var}

\end{quote}

Statements are either expressions, or variable declarations.  A
variable declaration must include an initial value.  Object variables are
declared with the \keyw{val} or \keyw{var} keyword; the former
indicates a final let binding, whereas the latter indicates a
assignable variable that can be updated.
State variables are declared with the \keyw{stateval} keyword. 

An optional type may be given for variable declarations.  If the type
is omitted for a \keyw{val} declaration, then it has the
structure of the initializing expression and the default permission
 for that structure. If no type is given for a \keyw{var} 
declaration, the variable is has type \keyw{dynamic}.

Statements evaluate to values, based on the expression in the
statement or the value of the initializer for the variable.  The last
statement in a sequence is used for the return value of a method or
the result of a block.

\subsection{Expressions}

\begin{quote}

\ntermdef{Expr}


\defspace \keyw{fn}   \opt{\nterm{MetaArgsSpec}}(\opt{\nterm{Args}})
              \opt{[\nterm{Args}]}
              => \nterm{Expr}

\defspace \nterm{Expr1}

\end{quote}

A first-class function includes standard polymorphic and parameter
argument declarations. The optional arguments surrounded by [] specify
environment variables that are captured by the lambda at its declaration.
The syntax of polymorphic parameters are described on
Section \ref{sec:polymorphic}. Polymorphic parameters specify which
data group permission a functions requires and allow the function to
be generic on which data groups it operates (similar to generic
methods on Java).
  
\begin{quote}

\ntermdef{Expr1}

\defspace \opt{\nterm{SimpleExpr} .} \nterm{Identifier} = \nterm{Expr}

\defspace \nterm{SimpleExpr} \texttt{<- }\nterm{State}

\defspace \nterm{SimpleExpr} \texttt{<{}<-} \nterm{State}

\defspace \keyw{atomic} \nterm{MetaArgs}  BlockExpr

\defspace \keyw{split} \nterm{MetaArgs}  BlockExpr

\defspace \keyw{unpack} BlockExpr

\defspace \keyw{match} ( \nterm{InfixExpr} ) \{ \seq{CaseClause} \}

\defspace \nterm{InfixExpr}

\end{quote}

The assignment form is for fields or for already-declared local
variables, which must have been declared using \keyw{var}. 

The state change operator \texttt{<-} modifies the object to the left of the
arrow as follows:

\begin{itemize}

\item
All tags on the right are added to the object.  Old tags are kept
unless they are inconsistent with the new tags, i.e. the old tag and
new tag are (transitively) different cases of the same state.

\item
All members that were declared in tags being removed, are removed from
the object.

\item
All members on the right are added to the object.  All old members on
the left that are not explicitly removed according to the bullet
above are retained.  

\item 
Futher details on the semantics of state change can be found in \citet{sunshine2011}.

\end{itemize}

The replacement operator \texttt{<{}<-} removes all tags and members from the object on the left and adds all tags and members of the state on the right.   

The type of either state change operations is \keyw{void}.

This paragraph provides a short overview of the \AE{}minium-specific
expressions. For a full definition of the semantics refer to
\cite{stork09:concurrency_by_default, stork10:uaeminium_spec}.  The
\keyw{atomic} expression provides a safe access environment to all
shared objects which belong to the data groups mentioned by the
\keyw{atomic} block. The \emph{MetaArgs} describes which data groups
the atomic block guarantees mutual exclusive access. Each
\emph{MetaArg} must refer to a valid data group (i.e., a concrete
declared data group or a group parameter).  The \keyw{split} executes
all statements of its body concurrently. To allow parallel access to
shared data the \keyw{split} block will split the declared data group
permissions into shared permissions, one for each statement.  
\keyw{unpack} is used to trade the group/access permission to the
specified object to gain access to the inner/nested groups declared
inside the object.

The \keyw{match} expression matches an input expression to one of
several cases using the \nterm{CaseClause} construct defined below.
The overall match expression evaluates to whatever value the
chosen case body evaluates to.

\begin{quote}

\ntermdef{CaseClause}

\defspace \keyw{case} \nterm{Pattern} \nterm{BlockExpr}

\defspace \keyw{default} \nterm{BlockExpr}

\ntermdef{Pattern}

\defspace \nterm{QualifiedIdentifier}


%\defspace \keyw{default}


\ntermdef{QualifiedIdentifier}

\defspace \nterm{Identifier} \seq{ . \nterm{Identifier}}

\end{quote}

The value is matched against each of the cases in order.  For the
first case that matches, the corresponding expression list is
evaluated.  If no pattern matches, an exception is
thrown.

A \keyw{case} tests the value's tags against
the \nterm{QualifiedIdentifier} given by the specified pattern.  The match succeeds if
one of the tags of the matched value is equal to the tag
\nterm{QualifiedIdentifier}, or if one of the tags of the matched value
was declared in a state that is a transitive \keyw{case of} the
\nterm{QualifiedIdentifier} specified.

The \nterm{QualifiedIdentifier}
must resolve to a state declared with the \keyw{state}
keyword; otherwise, an exception is
thrown. 

For the \keyw{default} case, the match always succeeds.  If
there is a \keyw{default} case, it must be the last one in the match
expression.



\begin{quote}

\ntermdef{InfixExpr}

\defspace \nterm{SimpleExpr}

\defspace \nterm{CastExpr}

\defspace \nterm{InfixExpr} \nterm{IdentifierOrOperator} \nterm{InfixExpr}


\ntermdef{IdentifierOrOperator}

\defspace \nterm{Identifier} \alt \nterm{Operator}


\ntermdef{CastExpr}

\defspace \nterm{SimpleExpr} \opt{\keyw{as} \nterm{Type}}

\end{quote}

The operators defined in Java have the same precedence in Plaid as
they do in Java, except the ternary operator and right shift operators, 
which are unsupported.  Identifiers as well as symbolic operators can be
used as infix operators; both are treated as method calls on the
object on the left of the operator.  Non-Java operators and
identifiers used as infix operators have a precedence above assignment
and state change, and below all other operators.

Cast expressions assert that a variable has a given type, and
also assert the relevant permission for that variable.  These casts
are trusted by the typechecker, but unchecked. A program that executes 
an invalid cast may fail at any point later in the program's 
execution.

\begin{quote}

\ntermdef{SimpleExpr}

\defspace \nterm{BlockExpr}

\defspace \keyw{new} \nterm{State}

\defspace \nterm{SimpleExpr2}

\end{quote}

The \keyw{new} statement creates an object initialized according to the
\nterm{State} specification given (defined below).

\begin{quote}

\ntermdef{BlockExpr}

\defspace \{ \opt{\nterm{StmtListSemi}} \}

\ntermdef{StmtListSemi}

\defspace \nterm{Stmt} \seq{ ; \nterm{Stmt}} \opt{;}

\end{quote}

Block expressions have a semicolon-separated list of statements, with
an optional semicolon at the end.  The statement list evaluates to the
value given by the last statement in the list.

\begin{quote}

\ntermdef{SimpleExpr2}

\defspace \nterm{SimpleExpr1}

\defspace \nterm{SimpleExpr2} \nterm{BlockExpr}

\end{quote}

To enable control structures with a natural, Java-like syntax, we allow
a function to be invoked by passing a block expression as an argument. The block expression in this 
case is treated as a zero-argument, anonymous lambda.For example, the \keyw{if} construct is defined in 
Plaid Standard Library so one can write \code{\keyw{if}(\emph{boolean value})\{...\};} in Plaid. One 
difference from Java syntax is that all expressions must be followed by a semi-colon, so \keyw{if} 
and \keyw{while} in Plaid must have trailing semi-colons unlike in Java.


\begin{quote}

\ntermdef{SimpleExpr1}

\defspace \nterm{Literal}

\defspace \nterm{Identifier}

\defspace \keyw{this}

\defspace ( \nterm{Expr} ) 

\defspace \nterm{SimpleExpr1} . \nterm{Identifier}

\defspace \nterm{SimpleExpr1} . \keyw{new}

\defspace \nterm{SimpleExpr1}  \opt{\nterm{MetaArgs}} \nterm{ArgumentExpr}

\ntermdef{ExprList}

\defspace \nterm{Expr} \seq{ , \nterm{Expr}}


\ntermdef{ArgumentExpr}


\defspace (  \opt{\nterm{ExprList}} ) 

\end{quote}

\keyw{this} represents the receiver of a method call as in Java and is 
bound in method bodies declared as members of states.  Unlike Java,
\keyw{this} is not bound in field initializers.

Expressions can appear within parentheses as a comma
separated list representing a tuple.

Java constructors can be invoked by calling \keyw{new}
on the Java class name.

Function and method invocation are handled uniformly by
supplying the arguments as a tuple.  Applications can be
chained, supporting currying.  Polymorphic arguments
are specified at each call site as well.

\section{Polymorphism}
\label{sec:polymorphic}

\begin{quote}

\ntermdef{MetaArgsSpec}

\defspace < \nterm{MetaArgSpec} \opt{, \nterm{MetaArgSpec}} >

\ntermdef{MetaArgSpec}

\defspace \keyw{group} \opt{\nterm{GroupPermission}} \nterm{Identifier}

\ntermdef{GroupPermission}

\defspace \keyw{exclusive}

\defspace \keyw{shared}

\defspace \keyw{protected}

\ntermdef{MetaArgs}

\defspace < \nterm{SimpleExpr1} \opt{, \nterm{SimpleExpr1}} >

\end{quote}

Plaid supports polymorphism for data groups\footnote{Extending
  polymorphism to types should be straightforward.} and uses angle
brackets to enclose polymorphic parameters and arguments (similar to
Java's generics). A \emph{MetaArgSpec} describes a single
 polymorphic formal parameter. At the moment Paid only supports only
group parameters. A group parameter consists of the 
keyword \keyw{group} to identify this parameter as group parameter, and optional
\emph{GroupPermission} (only for state declarations) and the
name of the parameter. For more information about data groups and
group parameters refer to \cite{stork09:concurrency_by_default,
  stork10:uaeminium_spec}.

\section{Declarations}

\begin{quote}
\ntermdef{Decl}

\defspace \seq{\nterm{ModifierOrDefaultPermission}} \keyw{state} \nterm{Identifier}  \opt{\nterm{MetaArgs}} \\
          \indent~~~~~~~~~~~~~~~~~~~\opt{\keyw{case} \keyw{of} \nterm{QualifiedIdentifier}  \opt{\nterm{MetaArgs}}}
          	\opt{\nterm{StateBinding}} \opt{;}

\defspace  \seq{\nterm{ModifierOrDefaultPermission}} \keyw{stateval} \nterm{Identifier}  \opt{\nterm{MetaArgs}} \\
           \indent~~~~~~~~~~~~~~~~~~~\opt{\nterm{StateBinding}} \opt{;}

\defspace \seq{\nterm{Modifier}} \nterm{MSpec} ;

\defspace \seq{\nterm{Modifier}} \nterm{MSpec} \nterm{BlockExpr}

\defspace \seq{\nterm{Modifier}} \nterm{FieldDecl} ;

\defspace \seq{\nterm{Modifier}} \nterm{GroupDecl} ;




\end{quote}

\keyw{state} and \keyw{stateval} declarations specify the implementation of a state,
as specified in the state definition. The \keyw{state} keyword means that this state is given its
own \textit{tag} that can be used to test whether objects are in that state.  Only states declared with \keyw{state} can be given in a pattern for a case in a \keyw{match} statement.

The \keyw{case} \keyw{of} keyword sequence assigns a superstate. States have 
all of the members of a superstate. Different cases of the same superstate 
are orthogonal; no object may ever be tagged with two cases of the same superstate.

The final two declarations are for method and field declarations.  The
method declaration has a method header
and an optional method body.  If the body is missing then
the method is abstract and must be filled in by sub-states or when the
state is instantiated.

Fields and state operators are discussed in more detail below.

\pagebreak

\begin{quote}

\ntermdef{StateBinding}

\defspace = \nterm{State}

\defspace \{ \seq{\nterm{Decl}} \}

\end{quote}

\begin{quote}

\ntermdef{State}

\defspace \nterm{StatePrim} \seq{\keyw{with} \nterm{StatePrim}} % left-associative

\ntermdef{StatePrim}

\defspace \{ \seq{\nterm{Decl}} \}

\defspace \nterm{SimpleExpr1} \opt{\{ \seq{\nterm{DecOrStateOp}} \}}

\defspace \keyw{freeze} \nterm{SimpleExpr1}

\ntermdef{DeclOrStateOp}

\defspace \nterm{Decl}

\defspace \nterm{StateOp}

\ntermdef{StateOp}

\defspace \keyw{remove} \nterm{Identifier} ;

\defspace \keyw{rename} \nterm{Identifier} \keyw{as} \nterm{Identifier} ;

\end{quote}

A \nterm{StateBinding} can either be an assignment to a \nterm{State} or
a list of declarations in curly braces.

A \nterm{State} is a composition of primitive states separated by the
\keyw{with} keyword.   The \nterm{StatePrim} category includes a list
of declarations, or an expression, that should evaluate to a state,
modified by a list of declarations and state operations (\nterm{DeclOrStateOp}).  
Expressions that evaluation to an object can be transformed into a primitive state
with the \keyw{freeze} keyword. 

Composition is in general symmetric, as in traits.  It is an error if two states are composed with a member in common. 
The conflict can be resolved manually using state operators \keyw{remove} and \keyw{rename}, which
respectively discard or change the name of a member in a state. 

\begin{quote}


%%%%%%%%%%%%%%%%%% Fields and Methods %%%%%%%%%%%%%%%%%%%%%

\ntermdef{GroupDecl}

\defspace \keyw{group} \nterm{Identifier} = \keyw{new} \keyw{group} ;

\ntermdef{FieldDecl}

\defspace \nterm{ConcreteFieldDecl}

\defspace \nterm{AbstractFieldDecl}

\ntermdef{ConcreteFieldDecl}

\defspace \opt{\nterm{Specifier}} \opt{Type} \nterm{Identifier} = \nterm{Expr}

\ntermdef{AbstractFieldDecl}

\defspace \nterm{Specifier} \nterm{Identifier} 

\defspace \opt{Specifier} \nterm{Type} \nterm{Identifier}

\ntermdef{Args}

\defspace \opt{\nterm{ArgSpec}} \nterm{Identifier} \seq{ , \opt{\nterm{ArgSpec}} \nterm{Identifier}}

\ntermdef{ArgSpec}

\defspace \nterm{Type} \opt{$>>$ \nterm{Type}}

\ntermdef{MSpec}

\defspace \keyw{method} \opt{\nterm{Type}} \nterm{IdentifierOrOperator} \opt{\nterm{MetaArgsSpec}} ( \opt{\nterm{Args}} )
          \opt{[ \nterm{Args} ]}

\end{quote}

Declaring a new data group is realized via \nterm{GroupDecl}, where identifier is the name of new 
data group. Data groups need to be immediately initialized with a new data group.$\!$\footnote{Future 
version might allow the declaration of data groups without initialization to realize data group 
parameters within structural types.} 

The \nterm{FieldDecl} form should be familiar from Java-like
languages.  If no expression is given then the field is abstract.  
When fields are first defined a specifier (\keyw{var} or \keyw{val})
must be given; later, when the field is overridden and given a concrete
value, the specifier may be omitted.
\keyw{var} fields are assignable, \keyw{val} fields
are not.

If a type is missing and an expression is given for a
  \keyw{val} field, then the type of the field is inferred from the
  expression as in variable declaration statements.  
  If the type is missing and either no expression is
  given or it is a \keyw{var} field, then the type is \keyw{dynamic}.

Method headers include specifications for their arguments.
Each argument specification includes the required type at the time of
the method call or function application.  If the parameter ends the
call with a different type this is indicated with a $>>$ and the
resulting type. If no resulting type is specified then it defaults to
the required type.  If no argument specification is given for
a variable, then its starting and ending type defaults to \keyw{dynamic}.

The method header \nterm{MSpec} has a standard format (similar to
functions).  As in function declarations, programmers may optionally
include types for any captured environment variables within square brackets. For methods
declared within states, the distinguished variable \keyw{this},
representing the receiver of the method, may appear in this list.  If
it does not, then the default specification for \keyw{this} is to have a required
and resulting type of the structure representing the state the method is defined in
with the default permission for that state as described below.

The name of an method can be an operator, which supports a simple form of \emph{operator overloading}. 
Operator methods are semantically equivalent to other methods. They dispatch only on the receiver, so 
\code{x + y} calls the \code{+} method defined on \code{x} with \code{y} as an argument.

\begin{quote}

%%%%%%%%%%%%%%%%%% Modifiers %%%%%%%%%%%%%%%%%%%%%

\ntermdef{Modifier}

\defspace \keyw{override}

\defspace \keyw{requires}

\ntermdef{DefaultPermission}

\defspace \keyw{immutable}

\ntermdef{ModifierOrDefaultPermission}

\defspace \nterm{Modifier}

\defspace \nterm{DefaultPermission}

\end{quote}


\keyw{override} indicates that a method overrides a function of the
same name during composition.

\keyw{requires} is similar to \keyw{abstract} in Java.  However,
things are more interesting in Plaid, because one can pass around an
object that has abstract/required members.  It is not necessary to
use the \keyw{requires} modifier in state definitions; one can simply
leave off the definition of a function.  \keyw{requires} is necessary
in types, however, to distinguish the presence vs. absence of a
member in that type.  Unlike in Java, methods may be called on an
object that has a required member, but only if the type given to the
method's receiver does not expect that member to be present.

A state with a default permission of \keyw{immutable} means that the permission of any fields,
local variables, or parameters declared to have the structure represented
by the state defaults to \keyw{immutable} when a permission
is not specified.  If the state is not give the default \keyw{immutable}
then the default permission is \keyw{unique}.


